NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PE and PG lipids

### 2021-01-05 manuscriptPGPE.tex
Small updates in the text

\begin{abstract}
% insert abstract here
  Chemistry of lipid headgroups, the water facing components of cell membranes, varies between organisms and
  organelles and regulate cell functions via lipid-protein interactions.
  Because lipid membranes are in liquid state in physiological conditions,
  the conformational ensembles of lipids and their dependence of headgroup chemistry are difficult to resolve experimentally.
  Here, we combine solid state NMR experiments and MD simulations from NMRlipids open collaboration to
  resolve conformational ensembles of PC, PS, PG and PE lipid headgroups in liquid lamellar phase in various biologically relevant conditions.
  Interpretation of NMR experiments using the best simulations from the NMRlipids project suggests that all lipid headgroups sample wide range of
  conformations in neutral and charged cell membranes, and that the differences in headgroup chemistry
  appears in the probability weights of different conformations. Together with the analysis of protein-bound-lipids
  from the protein data bank (PDB), this suggests that lipids can bind to proteins in wide range of conformations
  independently on the headgroup chemistry. Therefore, the specific lipid-protein interactions are likely driven by
  protein-lipid interactions, rather than conformational restrictions of lipids. Our results paves the way to
  comprehensive understanding of lipid-protein interaction energetics, and understanding complex biomolecular assemblies
  such as membrane proteins.

@@

\section{Methods}
\subsection{Experimental C--H bond order parameters}
The headgroup and glycerol backbone C--H bond order parameters of POPE and POPG
were measured using natural abundance $^{13}$C solid state NMR spectroscopy
as described previously \cite{ferreira13,ferreira16}.
The magnitudes of order parameters were determined from the chemical-shift resolved dipolar splittings
using a R-type Proton Detected Local Field (R-PDLF) experiment~\cite{dvinskikh04} and
the signs from S-DROSS experiments~\cite{gross97} combined with SIMPSON simulations \cite{bak00}.
%The experiments were done in a Bruker Avance III 400 spectrometer operating at a $^1$H Larmor frequency of 400.03 MHz.
%Magic angle spinning (MAS) of the sample was used at a frequency of 5.15 kHz (R-PDLF experiment) and 5 kHz (S-DROSS experiment).
The NMR experiments were identical as in our previous work~\cite{antila19}.
POPE and POPG powder were purchased from Avanti polar lipids \todo{Which isomer of POPG we have in the experiments?}.
The POPE experiments were recorded at 310~K and POPG experiments at 298~K, where the bilayers are in the liquid disordered phase \cite{marsh13}.

Glycerol backbone peaks from both lipids, and $\alpha$-carbon peak from POPE in the INEPT spectra 
were assigned based on previously measured POPC spectra~\cite{ferreira13}.
The $\beta$-carbon peak from POPE was assigned based on $^{13}$C chemical shift table for amines available
at \url{https://www.chem.wisc.edu/areas/reich/nmr/c13-data/cdata.htm}.
\todo{How were $\alpha$ and  $\gamma$-carbon peaks assigned in POPG?}.
The $\beta$-carbon peak from POPG overlapped with the g$_2$ peak from glycerol backbone
because their chemical environments are similar in POPG.
\todo{Details to be checked by Tiago}.

@@

Molecular dynamics simulation data were collected using
the NMRlipids Open Collaboration \cite{botan15}, which is running at
\url{nmrlipids.blogspot.fi} and \url{github.com/NMRlipids/NMRlipidsIVotherHGs}.
The simulated systems of pure PE and PG bilayers without additional ions
are listed in Tables \ref{systemsPE} and \ref{systemsPG},
and lipid mixtures with additional ions in Table~\ref{systemsMIX}.
Further simulation details are given in the SI, and
the simulation data are indexed in a
searchable database available at \url{www.nmrlipids.fi},
and in the NMRlipids/MATCH repository (\url{github.com/NMRlipids/MATCH}).

\subsection{Analysis of molecular dynamics simulation data}
The big data set of MD simulations was analysed in the NMRlipids databank manner.
Unique naming convention for lipid atoms in each force field was defined using the mapping files
and analysis for all simulations indexed in NMRlipids databank manner were performed
using python codes.

The C--H bond order parameters were calculated directly
from the carbon and hydrogen positions using the definition
\begin{equation}
S_{\rm CH}=\frac{1}{2}\langle 3\cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between the C--H bond and the membrane normal
(taken to align with $z$, with bilayer periodicity in the $xy$-plane).
Angular brackets denote average over all sampled configurations.
The order parameters were first calculated averaging over time separately
for each lipid in the system. The average and
the standard error of the mean were then calculated over different lipids.
Python programs that use the MDAnalysis library \cite{agrawal11,gowers16}
used for all atom simulations is available in Ref. \citenum{MATCHgit}
({\tt scripts/calcOrderParameters.py}). For united atom simulations, the trajectories
with hydrogens having ideal geometry were constructed first using either buildH program~\cite{buildH}
or ({\tt scratch/opAAUA\_prod.py}) in  Ref. \citenum{MATCHgit}, and the order parameters were
then calculated from these trajectories. This approach has been tested against trajectories
with explicit hydrogens and the deviations in order parameters are small \cite{buildH,piggot17}.\\
\todo{BuildH program is now cited with a direct link to the GitHub repo. I think that a release to Zenodo would be nice in the final publication.}\\
\todo{Maybe we should also shortly discuss here about the reasons for slight dependence of order parameter values on the method used to reconstruct hydrogens?}\\
The ion number density profiles were calculated using the {\tt gmx density} tool
of the Gromacs sofware package \cite{gromacsMANUAL}.

@@

\subsection{Analysis of lipid conformations bound to proteins}
Lipid structures from PDB (\url{http://www.rcsb.org/})
were searched using PDBe REST API (\url{www.ebi.ac.uk/pdbe/pdbe-rest-api}).
First, all PDB entries containing PC, PE, PG, or PS lipid headgroups were searched.
The ligand names to identify the lipids were:
PLC, PX4, 6PL, LIO, HGX, PC7, PC8, P1O, 6O8, XP5, EGY, PLD, SBM, HXG, and PCW for PC;
8PE, PTY, 3PE, PEH, PEF, 6OE, 6O9, 9PE, PEV, 46E, SBJ, L9Q, PEK, EPH, ZPE, 9TL, 9Y0, 6OU, LOP, and PEE for PE;
PGT, PGK, LHG, 44G, PGV, OZ2, D3D, PGW, DR9, P6L, PG8, H3T, and GOT for PG; and
PSF, PS6, Q3G, P5S, D39, PS2, 17F, and 8SP for PS.
All PDB structures containing these ligands were downloaded and the dihedral angles of
the first lipid in the latest version of the structures within each PDB code were calculated  
using MDTraj python library \cite{mcgibbon15}.
The used Jupyter notebook is available from the project's GitHub repository ({\it scripts/pdbSEARCH.ipynb}).

@@

\begin{figure*}[]
  \centering
   \includegraphics[width=18.0cm]{./Figs/figure2.eps}
   \caption{\label{structures}
     Results from CHARMM36 simulations demonstrating the differences in conformational ensembles betwee different lipids. 
     {\bf A)} Snapshots with overlayed C$_\beta$, C$_\alpha$ and O$_\alpha$ atoms and occurence of different conformations.
     {\bf B)} Headgroup and glycerol backbone region order parameters of different different lipids.
     {\bf C)} Distributions of heavy atom dihedral angles of different lipids from CHARMM36 simulations.
     {\bf D)} Distributions of P-N vector angle with respect to membrane normal.
  }
  \todo{This is a draft and requires quite a bit of polishing.
    More detailed discussion of this figure is in https://github.com/NMRLipids/NMRlipidsIVPEandPG/issues/9} \\
\end{figure*}

\subsection{Conformational ensembles of different lipid headgroups in bulk bilayer}

@@

Experimental order parameters of POPC, POPE, POPG and POPS glycerol backbones and
headgroups from this and previous studies are collected in Fig. \ref{HGorderParameters} D),
where signs determined from $^{13}$C NMR experiments are used also for the $^2$H NMR data from the literature. Overall agreement of order parameters determined by different authors and different techniques
for same lipid headgroup is very good in here and previous studies \cite{botan15,ollila16,antila19},
suggesting that the observed differences between lipid types arise from differences in
headgroup chemistry rather than inaccuracies in experiments. 
The most distinct order parameters are observed for PS headgroups,
for which the $\alpha$-carbon order parameter exhibits significant forking
and the $\beta$-carbon has more negative value than in other lipids.
On the other hand, the $\beta$-carbon order parameter of PG headgroup
has positive sign in contrast to all the other lipids.
Notably, this has not been observed in traditional $^2$H NMR experiments,
where only the absolute value of the order parameters are measured~\cite{wohlgemuth80,gally81,borle85}.
The glycerol backbone order parameters are similar for all the lipids, although they move slightly toward
positive values (closer to zero) in the order PC $<$ PE $<$ PS $<$ PG.
Essential differences between PC and PE headgroups are not observed.

\begin{figure*}[bt]
  \centering
  \includegraphics[width=18.0cm]{./Figs/HGorderparametersPCvsSURFchangeDIHEDRALS.eps}
  \caption{\label{changesWITHsurf}
    {\bf A)} Modulation of PC headgroup order parameters and P-N vector angle upon addition of cationic surfactant
    from CHARMM36 simulations compared with experimental data \cite{??}.
    {\bf B)} Changes in PC headgroup conformational ensembles upon increasing amount of positive charge in bilayer,
    characterized by the heavy atom dihedral distributions, from CHARMM36 simulations.
  }
\end{figure*}

To resolve the conformational ensembles of lipid headgroups,
we compared the headgroup and glycerol backbone C-H bond order parameters
from different MD simulation force fields to the experimental data. 
None of the force fields reproduced the order parameters
within experimental accuracy for any of the lipids analysed in this work
(Figs. \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG}) or in our previous studies \cite{botan15,antila19}.
Nevertheless, the CHARMM36 force field performs best for all the headgroups and
reproduces the distinct order parameters of PS and PG lipids (Fig. \ref{structures} B).
To understand the structural origin of these distinct order parameters,
we calculated the distributions of heavy atom dihedral angles from CHARMM36 simulations in figure \ref{structures} C).
Major differences between headgroups are observed only for the last two dihedrals
in the headgroup end,  O$_\alpha$-C$_\alpha$-C$_\beta$-N/C$_\gamma$ and P-O$_\alpha$-C$_\alpha$-C$_\beta$,
which prefer {\it gauche$^-$} conformations for PG and {\it gauche$^+$} for PS.
Rest of the dihedrals are similar between different lipids, with the exception of
small differences for PS lipids. Therefore we suggest that the main differences between
lipid headgroups leading to distinct order parameter occur in the choline part, while
also changes in phosphate region may contribute in PS lipids, which could also
explain the more rigid headgroup structures~\cite{browning80,buldt81}.
Furthermore, the angle between headgroup dipole and membrane normal
decreases in the order of PG $>$ PE  $>$ PC  $>$ PS (Fig. \ref{structures} B)).
However, the differences between PC and PE in P-O$_\alpha$-C$_\alpha$-C$_\beta$ dihedral
and P-N vector dipole may be an artefacts as the $\beta$-carbon order parameter in PC
is too negative the CHARMM36 force field, thereby not being equal to the order parameter
in PE as observed in experiments \cite{botan15}.

@@

Also the headgroup order parameters of charged lipids are affected by the binding of ions,
but their response is less well understood and more difficult to capture in MD simulations \cite{antila19,melcr20}.
In our simulations, the Lipid17 and Slipids force fields give the most realistic response of PG $\beta$-carbon to CaCl$_2$
in figure \ref{changesWITHCaClPG} even thought the calcium binding affinity was overestimated in these models. These models suggest that the upward tilting of the headgroup dipole is weaker in PG than in PC headgroup (\ref{changesWITHCaClPG}), but dihedral distributions are more sensitive to the bound ions in PG 
(Figs \ref{DIHSwithCAslipidsPOPG} and \ref{DIHSwithCAlipid17POPG}). However, none of the simulations captures the calcium binding affinity and conformational ensemble of PG lipids
simultaneously, experimental data to evaluate the response of $\alpha$-carbon order parameters to
the added calcium in PG is not available, and changes in dihedrals upon addition of cacium differ between
the best models (Figs \ref{DIHSwithCAslipidsPOPG} and \ref{DIHSwithCAlipid17POPG}).
Therefore, more accurate force fields are required for the detailed analysis of the interactions between ions and charged lipids, as concluded previously also for PS lipids \cite{antila19,melcr20}.

>>>>

\begin{abstract}
% insert abstract here
  Chemistry of lipid headgroups, the water facing components of cell membranes that regulate cell functions via lipid-protein interactions, varies between organisms and organelles. Because lipid membranes are in liquid state in physiological conditions, the conformational ensembles of lipid headgroups and their dependence on chemistry are difficult to resolve experimentally. Here, we combine solid state NMR experiments and MD simulations from NMRlipids open collaboration to resolve conformational ensembles of PC, PS, PG and PE lipid headgroups in liquid lamellar phase in various biologically relevant conditions. Interpretation of NMR experiments using the best simulations from the NMRlipids project suggests that all lipid headgroups sample wide range of conformations in neutral and charged cell membranes, and that the differences in headgroup chemistry appears in the probability weights of different conformations. Together with the analysis of protein-bound-lipids from the protein data bank (PDB), this suggests that lipids can bind to proteins in wide range of conformations independently on the headgroup chemistry. Therefore, the specific lipid-protein interactions are likely driven by protein-lipid interactions, rather than conformational restrictions of lipids. Our results paves the way to comprehensive understanding of lipid-protein interaction energetics, and understanding complex biomolecular assemblies such as membrane proteins.

@@

\section{Methods}
\subsection{Experimental C--H bond order parameters}
The headgroup and glycerol backbone C--H bond order parameters of POPE and POPG
were measured using natural abundance $^{13}$C solid state NMR spectroscopy
as described previously \cite{ferreira13,ferreira16}.
The magnitudes of order parameters were determined from the chemical-shift resolved dipolar splittings
using a R-type Proton Detected Local Field (R-PDLF) experiment~\cite{dvinskikh04} and
the signs from S-DROSS experiments~\cite{gross97} combined with SIMPSON simulations \cite{bak00}.
%The experiments were done in a Bruker Avance III 400 spectrometer operating at a $^1$H Larmor frequency of 400.03 MHz.
%Magic angle spinning (MAS) of the sample was used at a frequency of 5.15 kHz (R-PDLF experiment) and 5 kHz (S-DROSS experiment).
The NMR experiments were identical as in our previous work~\cite{antila19}.
POPE and POPG powder were purchased from Avanti polar lipids 
\todo{POPG is racemic in experiments. How much we should discuss this in relation with simulations?}.
The POPE experiments were recorded at 310~K and POPG experiments at 298~K, where the bilayers are in the liquid disordered phase \cite{marsh13}.

Glycerol backbone peaks from both lipids, and $\alpha$-carbon peak from POPE in the INEPT spectra 
were assigned based on previously measured POPC spectra~\cite{ferreira13}.
The $\beta$-carbon peak from POPE was assigned based on $^{13}$C chemical shift table for amines available
at \url{https://www.chem.wisc.edu/areas/reich/nmr/c13-data/cdata.htm}.
\todo{How were $\alpha$ and  $\gamma$-carbon peaks assigned in POPG?}.
The $\beta$-carbon peak from POPG overlapped with the g$_2$ peak from glycerol backbone
because their chemical environments are similar.
\todo{Details to be checked by Tiago}.

@@

Molecular dynamics simulation data were collected using
the NMRlipids Open Collaboration \cite{botan15}, which is running at
\url{nmrlipids.blogspot.fi} and \url{github.com/NMRlipids/NMRlipidsIVotherHGs}.
The simulated systems of pure PE and PG bilayers without additional ions
are listed in Tables \ref{systemsPE} and \ref{systemsPG},
and lipid mixtures with additional ions in Tables~\ref{systemsMIX} and~\ref{systemsMIX2}.
Further simulation details are given in the SI, and
the simulation data are indexed in a
searchable database available at \url{www.nmrlipids.fi},
and in the NMRlipids/MATCH repository (\url{github.com/NMRlipids/MATCH}).

\subsection{Analysis of molecular dynamics simulation data}
The large set of MD simulation data was analysed using the development version of NMRlipids databank.
Unique naming convention for lipid atoms in each force field was defined using the mapping files
and analysis for all simulations indexed in NMRlipids databank manner were performed
using python codes.

The C--H bond order parameters were calculated directly
from the carbon and hydrogen positions using the definition
\begin{equation}
S_{\rm CH}=\frac{1}{2}\langle 3\cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between the C--H bond and the membrane normal
(taken to align with $z$, with bilayer periodicity in the $xy$-plane).
Angular brackets denote average over all sampled configurations.
The order parameters were first calculated averaging over time separately
for each lipid in the system. The average and
the standard error of the mean were then calculated over different lipids.
Python programs that use the MDAnalysis library \cite{agrawal11,gowers16}
used for all atom simulations is available in Ref. \citenum{MATCHgit}
({\tt scripts/calcOrderParameters.py}). For united atom simulations, the trajectories
with hydrogens having ideal geometry were constructed first using either buildH program~\cite{buildH}
or ({\tt scratch/opAAUA\_prod.py}) in  Ref. \citenum{MATCHgit}, and the order parameters were
then calculated from these trajectories. This approach has been tested against trajectories
with explicit hydrogens and the deviations in order parameters are small \cite{buildH,piggot17}.\\
%\todo{BuildH program is now cited with a direct link to the GitHub repo. I think that a release to Zenodo would be nice in the final publication.}\\
The ion number density profiles were calculated using the {\tt gmx density} tool
of the Gromacs sofware package \cite{gromacsMANUAL}.

@@

\subsection{Analysis of lipid conformations bound to proteins}
Lipid structures from PDB (\url{http://www.rcsb.org/})
were searched using PDBe REST API (\url{www.ebi.ac.uk/pdbe/pdbe-rest-api}).
First, all PDB entries containing PC, PE, PG, or PS lipid headgroups were searched.
The ligand names to identify the lipids were:
PLC, PX4, 6PL, LIO, HGX, PC7, PC8, P1O, 6O8, XP5, EGY, PLD, SBM, HXG, and PCW for PC;
8PE, PTY, 3PE, PEH, PEF, 6OE, 6O9, 9PE, PEV, 46E, SBJ, L9Q, PEK, EPH, ZPE, 9TL, 9Y0, 6OU, LOP, and PEE for PE;
PGT, PGK, LHG, 44G, PGV, OZ2, D3D, PGW, DR9, P6L, PG8, H3T, and GOT for PG; and
PSF, PS6, Q3G, P5S, D39, PS2, 17F, and 8SP for PS.
All PDB structures containing these ligands were downloaded and the dihedral angles of the first lipid in the latest version of the structures within each PDB code were calculated using the MDTraj python library \cite{mcgibbon15}.
The used Jupyter notebook is available from the project's GitHub repository ({\it scripts/pdbSEARCH.ipynb}).

@@

\begin{figure*}[bt]
  \centering
   \includegraphics[width=18.0cm]{./Figs/figure2.eps}
   \caption{\label{structures}
     Results from CHARMM36 simulations demonstrating the differences in conformational ensembles betwee different lipids. 
     {\bf A)} Snapshots with overlayed C$_\beta$, C$_\alpha$ and O$_\alpha$ atoms and occurence of different conformations.
     {\bf B)} Headgroup and glycerol backbone region order parameters of different different lipids.
     {\bf C)} Distributions of heavy atom dihedral angles of different lipids from CHARMM36 simulations.
     {\bf D)} Distributions of P-N vector angle with respect to membrane normal.
  }
  \todo{This is a draft and requires quite a bit of polishing.
    More detailed discussion of this figure is in https://github.com/NMRLipids/NMRlipidsIVPEandPG/issues/9} \\
\end{figure*}

\subsection{Differences between lipid headgroups in bulk bilayer from $^{13}$C NMR experiments}

@@

Experimental order parameters of POPC, POPE, POPG and POPS glycerol backbones and headgroups from this and previous studies are collected in Fig. \ref{HGorderParameters} D), where signs determined from $^{13}$C NMR experiments are used also for the $^2$H NMR data from the literature. Overall agreement of order parameters determined by different authors and different techniques for same lipid headgroup is very good in here and previous studies \cite{botan15,ollila16,antila19},
suggesting that the observed differences between lipid types arise from differences in headgroup chemistry rather than inaccuracies in experiments, or differences in acyl or experimental conditions. The most distinct order parameters are observed for PS headgroups, for which the $\alpha$-carbon order parameter exhibits significant forking and the $\beta$-carbon has more negative value than in other lipids. On the other hand, the $\beta$-carbon order parameter of PG headgroup has positive sign in contrast to all the other lipids. Notably, this has not been observed in traditional $^2$H NMR experiments, where only the absolute value of the order parameters are measured~\cite{wohlgemuth80,gally81,borle85}. The glycerol backbone order parameters are similar for all the lipids, although they move slightly toward positive values (closer to zero) in the order PC $<$ PE $<$ PS $<$ PG. Essential differences between PC and PE headgroups are not observed.

\subsection{Conformational ensembles of different lipid headgroups in bulk bilayer from MD simulations}

\begin{figure*}[bt]
  \centering
  \includegraphics[width=18.0cm]{./Figs/HGorderparametersPCvsSURFchangeDIHEDRALS.eps}
  \caption{\label{changesWITHsurf}
    {\bf A)} Modulation of PC headgroup order parameters and P-N vector angle upon addition of cationic surfactant
    from CHARMM36 simulations compared with the experimental data \cite{scherer89}.
    {\bf B)} Changes in PC headgroup conformational ensembles upon increasing the amount of positive charge in bilayer,
    characterized by the heavy atom dihedral distributions, from CHARMM36 simulations.
  }
\end{figure*}

To resolve the conformational ensembles of lipid headgroups, we compared the headgroup and glycerol backbone C-H bond order parameters from different MD simulation force fields to the experimental data. None of the force fields reproduced the order parameters within experimental accuracy for any of the lipids analysed in this work (Figs. \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG}) or in our previous studies \cite{botan15,antila19}. Nevertheless, the CHARMM36 force field performs best for all the headgroups and reproduces the experimentally observed distinct order parameters of PS and PG lipids (Fig. \ref{structures} B). To understand the structural origin of these distinct order parameters, we calculated the distributions of heavy atom dihedral angles from CHARMM36 simulations in figure \ref{structures} C). Major differences between headgroups are observed only for the last two dihedrals in the headgroup end,  O$_\alpha$-C$_\alpha$-C$_\beta$-N/C$_\gamma$ and P-O$_\alpha$-C$_\alpha$-C$_\beta$, which prefer {\it gauche$^-$} conformations for PG and {\it gauche$^+$} for PS. Rest of the dihedrals are similar between different lipids, with the exception of small differences for PS lipids. Therefore we suggest that the main differences between lipid headgroups leading to distinct order parameter occur in the choline part, while also changes in phosphate region may contribute in PS lipids, which could also explain the more rigid headgroup structures~\cite{browning80,buldt81}.
Furthermore, the angle between headgroup dipole and membrane normal decreases in the order of PG $>$ PE  $>$ PC  $>$ PS (Fig. \ref{structures} B)). However, the differences between PC and PE in P-O$_\alpha$-C$_\alpha$-C$_\beta$ dihedral
and P-N vector dipole may be artificial as the $\beta$-carbon order parameter in PC is too negative in the CHARMM36 force field, thereby not being equal to the order parameter in PE as observed in experiments \cite{botan15}.

@@

Also the headgroup order parameters of charged lipids are affected by the binding of ions, but their response is less well understood and more difficult to capture in MD simulations \cite{antila19,melcr20}. In our simulations, the Lipid17 and Slipids force fields give the most realistic response of PG $\beta$-carbon to CaCl$_2$ in figure \ref{changesWITHCaClPG} even thought the calcium binding affinity was overestimated in these models. These models suggest that the upward tilting of the headgroup dipole is weaker in PG than in PC headgroup (\ref{changesWITHCaClPG}), but dihedral distributions are more sensitive to the bound ions in PG (Figs \ref{DIHSwithCAslipidsPOPG} and \ref{DIHSwithCAlipid17POPG}). However, none of the simulations captures the calcium binding affinity and conformational ensemble of PG lipids simultaneously, experimental data to evaluate the response of $\alpha$-carbon order parameters to the added calcium in PG is not available, and the changes in dihedrals upon addition of cacium differ between the best models (Figs \ref{DIHSwithCAslipidsPOPG} and \ref{DIHSwithCAlipid17POPG}). Therefore, more accurate force fields are required for the detailed analysis of the interactions between ions and charged lipids, as concluded previously also for PS lipids \cite{antila19,melcr20}.
